%% This is the preface for my UBC PhD Dissertation
%% The parent document is called thesis.tex

\chapter{Preface}
Chapter~\ref{ch:fall_sim_design} has been published as a design innovation paper~\citep{gilchrist_development_2013}.
I was responsible for conceptual design of the impact apparatus, literature review in support of the design, construction of the device, and validation of its performance.
Dr.\ Cripton and Dr.\ Guy provided guidance, supervision and edited the manuscript.

Chapter~\ref{ch:behave_fail} has been submitted for publication as an original paper~\citep{gilchrist_mechanics_2013}.
I was lead investigator on this research, however, many other researchers contributed time and data to the investigation.
The paper included new analyses of data from previous researchers.
The specimens and data affiliated with the \textit{slow} group were tested by myself and Dr.\ Nishiyama. Dr.\ Nishiyama prepared the specimens for testing and performed pre-test imaging on the specimens.
Dr.\ Boyd provided supervision and funding for Dr.\ Nishiyama and proof reading of the paper.
The specimens and data affiliated with the \textit{fast} group were provided by Dr.\ de Bakker, who was supervised and funded by Dr.\ Oxland.
I was responsible for analysing the data associated with these specimens.
I collected, prepared and tested and analysed the \textit{fall} group specimens.
All statistical analyses were carried out by me.
Dr.\ Cripton and Dr.\ Guy provided supervision and edited the manuscript.

I was lead investigator for the project described in Chapter~\ref{ch:fracture}.
I was responsible for design of the tests, strain field analyses, video observations of fracture initiations, statistical analysis and writing of the manuscript.
Dr.\ Nishiyama provided the quasi-static specimens and aided in testing them, and Dr.\ Boyd supervised and funded him.
Dr.\ Guy assisted in sketching the fracture lines and classified the fractures into clinical types.
He also assisted with manuscript preparation and editing.
Dr.\ Cripton advised me and assisted in manuscript preparation and editing.

I was lead investigator and study designer for the work presented in Appendix~\ref{ch:version}, which will be submitted for publication.
I performed all digital scans, and developed the methods for measuring the femoral version and the linea aspera angle and helped develop the posterior neck angle measurement.
Michael Cancilla developed the posterior neck angle measurement and performed inter-observer reliability readings for other measures.
Dr.\ Guy assisted in developing the posterior neck angle measurement, provided supervision and edited the manuscript.
Dr.\ Cripton provided supervision and edited the manuscript generated from the chapter that is currently in submission.

I was lead investigator for the work presented in Appendix~\ref{ch:dvc}.
I developed all software and apparatus, and performed all validation testing.
Dr.\ Liebschner provided images for the work presented in \S\ref{sec:dvc_results_human}.
Drs.\ Cripton and Guy provided funding, supervision and editing of posters and presentations associated with this work.

The work in Chapters~\ref{ch:behave_fail} and~\ref{ch:fracture} were approved by the \ac{ubc} Clinical Ethics review board (certificate number: H06-70337).
The work in Appendix~\ref{ch:version} was approved by the \ac{ubc} Department of Anatomy (approval number: W0105).