%% This is the acknowledgments for my UBC PhD Dissertation
%% The parent document is called thesis.tex

\chapter{Acknowledgments}
The work in this thesis would not have been possible without support and assistance from a number of people and institutions.

I would like to start by thanking my supervisors and advising committee for their guidance and help throughout the process.
As co-supervisors, Drs.~Peter Cripton and Pierre Guy provided invaluable support in the completion of this work and in my training as a researcher.
As members of my advising committee, Drs.~Rizhi Wang, Dave Wilson and Thomas Oxland provided support and guidance throughout the duration of my studies.
I am very appreciative of the time and effort that these researchers have given to my personal and professional development.

Secondly, I would like to acknowledge my funding sources.
Four years of stipend and research funding were provided by the \ac{bridge} fellowship program.
The \ac{bridge} program also provided me with numerous learning and development opportunities, all of which were beneficial to my professional development and enjoyable on a personal level.
Funding for purchasing specimens and other laboratory equipment was provided by an \ac{oref} grant from the Department of Orthopaedics at the University of British Columbia.
Additional funding for laboratory space and other laboratory equipment was provided by a \acf{nserc} operating grant.
Finally, the facilities and institutes in which the tests and analysis were conducted (namely, the Blusson Spinal Cord Centre with \acs{icord} and the Robert H.\ N.\ Ho Building with the \acs{chhm}) were funded by the Canadian Foundation for Innovation.

There were also a number of faculty and students who assisted me both technically and emotionally over the course of my studies.
Angela Melnyk, Jenn Douglas and Oscar Ariza of the \ac{oibg}, Dr.\ Ben Helgason of ETH-Z\"{u}rich and Dr.\ Danmei Liu of the \ac{chhm} were all instrumental in completion of these studies.

Finally, I would like to thank my wife, Claire, my family, and all of my friends for their support over the years.
